---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bssdm: Biosecurity Species Distribution Modelling

<!-- badges: start -->
[![Last commit](https://img.shields.io/github/last-commit/cebra-analytics/bssdm.svg)](https://github.com/cebra-analytics/bssdm/commits/main)
<!-- badges: end -->

The *bssdm* package provides Species Distribution Modelling (SDM) implementations for two presence-only methods for predicting the spatial distribution environmental suitability for exotic pests, diseases, and other biosecurity threats, utilising spatial species occurrence records and their corresponding environmental variables. The two methods are:

1. *Range bagging* (Drake, 2015)
1. *Climatch* (ABARES, 2020)

## Installation

You can install the latest version of *bssdm* from [GitHub](https://github.com/) with:

``` r
remotes::install_github("cebra-analytics/bssdm")
```

## Example

The following example generates the environmental suitability for a Hawkweed species (*Hieracium pilosella*), an exotic weed for Australia, using both the *bssdm* package *Range bagging* and *Climatch* method implementations.

### Step 1: Obtain environmental variables

The SDM requires spatial environmental data in the form of *GeoTIFF* raster layers for the area of interest, encapsulating the species occurrence records (to build the model), as well as locations where the predicted environmental suitability is desired. Here we will use a selection of global climate data from *WorldClim* (Fick & Hijmans, 2017; <http://www.worldclim.org>).

```{r, message = FALSE, fig.align = "center"}
# Climate WorldClim (BIOCLIM) data
# BIO02: Mean Diurnal Range
# BIO05; Max Temperature of Warmest Month
# BIO11: Mean Temperature of Coldest Quarter
# BIO12: Annual Precipitation
# BIO14: Precipitation of Driest Month
# BIO15: Precipitation Seasonality (Coefficient of Variation)
climate_rast <- terra::rast(sprintf("data/bioclim_10m/wc2.1bio%02d.tif",
                                    c(2,5,11,12,14,15)))
climate_rast
```

### Step 2: Obtain species occurrence records

The SDM requires species occurrence records to be specified in a table with latitude and longitude coordinates using the WGS84 coordinate reference system (CRS). Here we will use global Hawkweed (*Hieracium pilosella*) occurrences downloaded from Global Biodiversity Information Facility (GBIF, 2026).

It is recommended to firstly 'clean' occurrence records to remove duplicates and incomplete or incorrect records via a tool such as *CoordinateCleaner* (Zizka et al., 2019). We will use occurrence records that were cleaned via *CoordinateCleaner*.

```{r, message = FALSE, fig.align = "center"}
# Load cleaned Hawkweed (Hieracium pilosella) occurrences
occurrences_cleaned <- read.csv("data/hawkweed_occurrences_cleaned.csv")
head(occurrences_cleaned)
# Plot BIOCLIM BIO02 & occurrences
terra::plot(climate_rast[[1]], colNA = "grey",
            main = "BIOCLIM Mean Diurnal Range & Hawkweed occurrences (red)",
            xlab = "Longitude (degrees)", ylab = "Latitude (degrees)")
terra::plot(terra::vect(occurrences_cleaned, crs = "EPSG:4326"),
            col = "red", pch = 20, alpha = 0.5, add = TRUE)
```

### Step 3: Run the SDM

To run a SDM we first build a model, then use it to predict the suitability for the area of interest using climate data with matching variables. Although this climate data may differ in its extent, CRS, resolution, or time frame (e.g. past or future climate), here we reuse the climate data used to build the model. We will build models and predict suitability for both our *Range bagging* and *Climatch* SDM methods.

#### Range bagging SDM

```{r, message = FALSE, fig.align = "center"}
# Run Range bagging SDM
rangebag_model <- bssdm::rangebag(climate_rast, occurrences_cleaned)
rangebag_output <- predict(rangebag_model, climate_rast, raw_output = FALSE)
# Plot the Range bagging SDM predicted climate suitability
terra::plot(rangebag_output, colNA = "grey",
            main = "Range bagging SDM predicted climate suitability",
            xlab = "Longitude (degrees)", ylab = "Latitude (degrees)")
```

#### Climatch SDM

```{r, message = FALSE, fig.align = "center"}
# Run Climatch SDM
climatch_model <- bssdm::climatch(climate_rast, occurrences_cleaned)
climatch_output <- predict(climatch_model, climate_rast, raw_output = FALSE)
# Plot the Climatch SDM predicted climate suitability
terra::plot(climatch_output, colNA = "grey",
            main = "Climatch SDM predicted climate suitability",
            xlab = "Longitude (degrees)", ylab = "Latitude (degrees)")
```

## References

ABARES (2020). 'Climatch v2.0 User Manual'. Canberra. <https://climatch.cp1.agriculture.gov.au/> Accessed: November 2021.

Drake, J. M. (2015). 'Range bagging: a new method for ecological niche modelling from presence-only data'. *Journal of the Royal Society Interface*, 12(107), 20150086. <doi:10.1098/rsif.2015.0086>

Fick, S. E., Hijmans, R. J. (2017). 'WorldClim 2: new 1-km spatial resolution climate surfaces for global land areas'. *International Journal of Climatology*, 37, 4302–4315. <doi:10.1002/joc.5086>

GBIF.org (04 May 2026) 'GBIF Occurrence Download'. <doi:10.15468/dl.q6q6fk>

Zizka, A., Silvestro, D., Andermann, T., Azevedo, J., Duarte Ritter, C., Edler, D., Farooq, H., Herdean, A., Ariza, M., Scharn, R., Svanteson, S., Wengstrom, N., Zizka, V., & Antonelli, A. (2019). 'CoordinateCleaner: standardized cleaning of occurrence records from biological collection databases.' *Methods in Ecology and Evolution*, 7. <doi:10.1111/2041-210X.13152>, R package version 3.0.1, <https://github.com/ropensci/CoordinateCleaner>.
